Edge sign prediction in general graphs and Correlation Clustering

### 2017-10-24 edge/model.tex
take Marc feedback on the edge model into account
- split into more subsection
- add an example
- describe and name the methods earlier
- add a paragraph comparing all the methods
- make it more clear there will be only synthetic data

\section{Learning problem}
\label{sec:edge_problem}

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm\stodo{or bounded?}. As we will discuss, this
constraint can be seen as a way to couple the different dimensions of the profiles%
\stodo{don't remember what I meant here}.

@@

We assume in the following that both the profiles and the directions are vectors with components in
$[-1, 1]$. A natural choice for the operator $s$ is then the difference between the profiles, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_{uv}$, we expect
${x_u}^T w_{uv}$ to be close to ${x_v}^T w_{uv}$, as measured by $|{x_u}^T w_{uv} - {x_v}^T
w_{uv}|$. Because we prefer to work with smooth functions, we define the goodness to be $g(s_{uv},
w_{uv}) = \left({s_{uv}}^T w_{uv} \right)^2$, and we want to minimize it with respect to $w_{uv}$.
When on the \ith{} dimension ${x_u}_{;i} - {x_v}_{;i}$ is close to zero, the minimization objective
and the fact that $w_{uv}$ is
unit-norm impose that ${w_{uv}}_{;i}$ should be large. On the other hand, with this formulation,
there is no way to set ${w_{uv}}_{;i}$ in order to \enquote{highlight} the fact that
${x_u}_{;i}$ and ${x_v}_{;i}$ are widely different.
Whereas there might be situations where this is not a problem, as we discussed in the introduction,
here we seek to model heterophilic interactions, at least along a subset of dimensions. Therefore, in
the following, we will adopt a refined approach.

@@

Based on these specific definitions of $s$ and $g$, we now formally write down the problem of
finding and assigning the most informative directions to every edge of $G$
\begin{problem}[]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ unit-norm directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dsphere$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{w_1,\ldots,w_k \in \dsphere}
    \sum_{u,v \in E} \max_{\ell \in \rangesk} g(s_{uv}, w_{\ell}) =
    \sum_{u,v \in E} \max_{\ell \in \rangesk} \left(x_u \circ x_v\right)^T w_{\ell} = f(k)
  \end{equation}
  where $s_{uv} = x_u \circ x_v$ and $g(s_{uv}, w_{\ell}) = {s_{uv}}^T w_{\ell}$.
\end{problem}

@@

Here we show why \autoref{p:edge_full} can model signed graphs as a special case, namely with $k=2$
and specifically crafted profiles.
Remember the learning
bias we introduced in the previous chapter \vpageref{text:cc_new_bias}: every node belongs to
exactly one of the $K$ clusters available, is connected positively with the other nodes belonging to
the same cluster, and is connected negatively with the nodes belonging to every other clusters. Now
let the dimension of profiles be $d=K$ and the number of directions $k=2$. Moreover, we define the
two direction vectors as $w_1 = \frac{\onev}{\sqrt{d}} = w^+$ and $w_2 = -w_+ = w_-$, where
$\onev$ is the vector of dimension $d$ with only $1$ coordinates. Finally, for a node $u$ in
cluster $i$, we assume its profile $x_u$ is set as follow: all component are equal to $-b$ except
for the \ith{} one, which is equal to $a$, where $a$ and $b$ are real constants we define now.

First, because $x_u$ is a unit vector, we have that $a^2 + (K-1)b^2=1$, from which $a =
\sqrt{1-(K-1)b^2}$ whenever $1-(K-1)b^2 \geq 0$, or equivalently $b \leq \frac{1}{\sqrt{K-1}}$.
Next, we look at $\left(x_u \circ x_v\right)^T \onev = {s_{uv}}^T \onev = \frac{1}{\sqrt{d}}
\sum_{i=1}^d  {s_{uv}}_{;i}$ for any edge $(u,v) \in E$. If $u$ and $v$ are in the same cluster, then
$\sum_{i=1}^d  {s_{uv}}_{;i} = a^2 + (K-1)b^2 = 1$, implying that ${s_{uv}}^T w_+ = \frac{1}{d} >
{s_{uv}}^T w_- = -\frac{1}{d}$. Otherwise, $\sum_{i=1}^d  {s_{uv}}_{;i} = (K-2)b^2 - 2ab = (K-2)b^2 -
2b\sqrt{1-(K-1)b^2} = f_K(b)$. By computing the derivative of $f_K$ with respect to $b$, we find
that it reaches a minimum of $-\frac{1}{K-1}$ at $b = \frac{1}{\sqrt{K(K-1)}} \leq
\frac{1}{\sqrt{K-1}}$. In this case, ${s_{uv}}^T w_+ = -\frac{1}{(K-1) \sqrt{d}} < {s_{uv}}^T w_- =
\frac{1}{(K-1) \sqrt{d}}$. To maximize $\sum_{(u,v)\in E} \max_{1 \leq \ell \leq d} w_\ell^T
s_{uv}$, we therefore assign $w_+$ to edges whose both endpoints belong to the same cluster, and
$w_-$ to edges joining two different clusters.

@@

\bigskip

At the other extreme of having only $k=2$ directions, when $k=|E|$, the topology of the graph
becomes irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} f(k)$ to increase as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers. To retain a fair level of interpretability, we deem appropriate to have $k$ not
too large.

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level. 
Recall that we consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. One way to integrate that fact in
our optimization formulation is to assume that profiles are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real coefficients and $b_u
\in \Rbb^d$. Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it consume a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term 
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} = 
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}
from \eqref{eq:edge_full}.

We can also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
certain value of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{u,v}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.} and $c_u$ be its $u^{th}$ row. Then $Yc_u$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of all $Yc_u$
(\ie its number of non-zero component) to be upper bounded by $k_\mathrm{local}$. As previously,
this is not convex but we can relax this constraint, as such sparsity inducing problem have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership vector, that
is we set $y_{uv,i} = \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_i)}$,
and we replace the $\ell_0$ norm by a $\ell_1$ or even a $k$-support norm~\autocite{KsupportNorm12}.
These two constraints can be combined, so that the profiles are close to a linear combination of a
small number of directions.

@@

A simple and natural baseline is to cluster the set of all similarity edge vectors $
\theset{s_{uv}}{(u,v) \in E}$ using the $k$-means algorithm, where $k$ is the number of directions
set in \autoref{p:edge_full}. Formally, given the graph $G=(V,E)$ and the node profiles $X$ as
input, we order the edges from $e_1$ to $e_m$ and build the matrix $S \in \Rbb^{m\times d}$ whose
\ith{} row is the similarity between the profiles of the endpoints of $e_i=(u,v)$, that is $s_{uv} =
x_u \circ x_v$. Running $k$-means will thus partition $E$ into $k$ sets of edges, call them $E_1,
\ldots, E_k$. The first drawback of that simplicity is that it does not solve \autoref{p:edge_full}.
Indeed, the partition of $E$ does not provide the set of directions $\mathcal{D}_k = \{w_1, w_2,
\ldots, w_k\} \in \dsphere$ we are looking for. A natural way of obtaining directions from that
partition is to set $w_\ell$ to be the normalized cluster center of $E_\ell$. Note however that
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its score ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them, that is as previously, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number of iterations would be needed.

\subsection{Vector optimization}
\label{sub:edge_vector}

The difficulty in optimizing directly \eqref{eq:edge_full} stem from the $\max$ operator, which we
replace here by a convex relaxation. Indeed, one can check that given a set of numbers $S=\{a_1,
a_2, \ldots, a_{|S|}\}$ and real number $\beta > 0$, we have
\begin{equation*}
  \max_{a_i \in S} a_i = \lim_{\beta \rightarrow + \infty} \frac{1}{\beta}
  \log\left( \sum_{i=1}^{|S|} \exp{\beta a_i} \right)
\end{equation*}
Using this fact, and averaging the two components of our objective function, we can rewrite
\eqref{eq:edge_full} as:
\begin{equation}
  \label{eq:edge_soft}
  \argmax_{{\substack{w_1,\ldots,w_k \in \dsphere,\\
  a_{uv}\in\Rbb,\, b_u \in \Rbb^d}}}
  \frac{1}{|E|\beta} \sum_{u,v\in E}
  \log\Bigl(\sum_{\ell=1}^k\exp\left(\beta {s_{uv}}^T w_\ell \right)\Bigr)
  - \frac{1}{|V|}\sum_{u\in V}
  \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2
\end{equation}
Note that in second term, because we do not know the edge assignment, we have to decide which of the
$w_1, \ldots, w_k$ is $w_{uv}$ equal to. According to our definition of the goodness function, we
want to let $w_{uv} = \argmax_{w_\ell \in \mathcal{D}_k} {s_{uv}}^T w_\ell$. However, the $\argmax$
function is not convex and we thus use the same relaxation as before. Namely, by stacking all the
vectors of $\mathcal{D}_k$ into a single vector $W$ of size $k\times d$, the relaxation is:
$$ \begin{array}[t]{lrcl}
  \mathcal{A} : & \Rbb^m & \longrightarrow & \Rbb^d \\
      & (w_1,\ldots,w_k) & \longmapsto &
  \sum_{i=1}^k \frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_i)} w_i
\end{array} $$
This function has an explicit derivative, but it is rather costly to compute and in practice we
found automatic differentiation~\autocite{autograd15} to be more efficient. While this modified
objective function is now jointly convex in all its variables, the set $\dsphere$ is not. To ease
the optimization procedure, we thus only impose the directions $w_\ell$ to lie within the unit ball
and expect them to have norm close to one in order to maximize the dot product in the ${s_{uv}}^T
w_\ell$ term.

>>>>

\section{Attributed graphs and problem definition}
\label{sec:edge_problem}

We start by introducing some notations and terminology about edge types in node-attributed graphs.
Then we present two options to model the interactions between nodes and the role of attributes in
\enquote{explaining} edges (\autoref{sub:edge_setting}). Choosing of these two options leads us to
formulate the \ecp{} problem in \autoref{sub:edge_constraints}. We then describe additional
constraints to take into account the topology of the graph. Finally we elaborate on the relation
between \ecp{} and \esp{} in signed graphs.

\subsection{Setting and modelling}
\label{sub:edge_setting}

As in the previous chapter, we are given an undirected, unweighted graph $G=(V,E)$. This
time though, we also have side information in addition to the graph topology. Namely, each node $u$
is associated with a $d$-dimensional feature vector $x_u \in \Rbb^d$. As an example, in a social
network, $x_u$ would be the profile of user $u$, describing $u$'s demographics and preferences. We
are also given a combination operator $s$ between feature vectors, such that $s(x_u, x_v) = s_{uv}
\in \Rbb^d$. Finally, we consider distinguished feature vectors $w_{uv}$,
along which the combination $s_{uv}$ between two nodes $u$ and $v$ is evaluated. We call such
$w_{uv}$ \emph{directions}, and constrain them to have unit norm.
While being exactly of norm one is not primordial, the role of bounding the norm of the directions
(\ie{} having $\sum_{i=1}^d {w_{uv}}_{;i}^2 \leq B^2$) is to tie the $d$ dimensions together.
Without such a constraint, we could simply study each dimension separately, whereas this bound
adds a coupling across different attributes, since increasing the weight of an attribute mechanically
affects the others.

@@

As written in the introduction, because we are interested in (partial) heterophilic links, we assume
in the following that both the profiles and the directions are vectors with components in $[-1, 1]$.
Let us take the example of encoding the gender of a user in a social network, assuming there is only
two values possible: male and female. Traditionally, this attribute would be set to $1$ if the user
is female and $0$ otherwise. Instead, we use $+1$ and $-1$, to clearly emphasize the difference
between the two values. The same is valid for continuous variables. For instance, an attribute could
range from $-1$ to $+1$ to represent the ideological stance of a user on the conservative/liberal
scale~\autocite{newsBias15}.\footnote{Note of course that in both cases, negative values do not
represent a judgement but are merely a reversible mathematical convenience.} This applies even to
non diverging positive quantities such as age, which can be projected to $[-1, 1]$ using
pre-processing such as standardization or $z$-score.

A natural choice for the operator $s$ is then the difference between the profiles, that is
$s_{uv} = x_u - x_v$. In that spirit, if $u$ and $v$ are close along a direction $w_{uv}$, we expect
${x_u}^T w_{uv}$ to be close to ${x_v}^T w_{uv}$, as measured by $|{x_u}^T w_{uv} - {x_v}^T
w_{uv}|$. Because we prefer to work with smooth functions, we define the goodness to be $g(s_{uv},
w_{uv}) = \left({s_{uv}}^T w_{uv} \right)^2$, and we want to minimize it with respect to $w_{uv}$.
When, on the \ith{} dimension, ${x_u}_{;i} - {x_v}_{;i}$ is close to zero, the minimization objective
and the fact that $w_{uv}$ is
unit-norm impose that ${w_{uv}}_{;i}$ should be large. On the other hand, with this formulation,
there is no way to set ${w_{uv}}_{;i}$ in order to \enquote{highlight} the fact that
${x_u}_{;i}$ and ${x_v}_{;i}$ are widely different.
There might be situations where this is not a problem. But again,
here we seek to model heterophilic interactions, at least along a subset of dimensions. Therefore,
we now introduce a more flexible approach.

@@

This second choice of $s$ and $g$ is indeed more flexible, for it encompasses all cases of
homophily, indifference and heterophily. Furthermore, since the profiles are given, the resulting
$s_{uv}$ can be seen as constants, meaning that $g$ is linear and therefore interpretable. Finally,
it also supports categorical data. We shall demonstrate all these features on a small scale example,
but we first define formally our problem.

\subsection{Learning problem and additional constraints}
\label{sub:edge_constraints}

\paragraph{Problem definition}

With working definitions of $s$ and $g$ at hand, we now formally write down the problem of
finding and assigning the most informative directions to every edge of $G$
\begin{problem}[\ecp{}]
  \label{p:edge_full}
  Given a graph $G=(V, E)$, node profiles $X\in [-1, 1]^{n\times d}$ and an integer $k \in \Nbb$,
  find a set of $k$ unit-norm directions $\mathcal{D}_k = \{w_1, w_2, \ldots, w_k\} \in \dsphere$
  and associate to every edge of $E$ the direction with the maximal goodness. Formally, solve
  \begin{equation}
    \label{eq:edge_full}
    \argmax_{w_1,\ldots,w_k \in \dsphere}
    \sum_{u,v \in E} \max_{\ell \in \rangesk} g(s_{uv}, w_{\ell})
  \end{equation}
  where $s_{uv} = x_u \circ x_v$ and $g(s_{uv}, w_{\ell}) = {s_{uv}}^T w_{\ell}$.
\end{problem}

\paragraph{Example}

In \autoref{fig:edge_exe}, we present a small instance of the \ecp{} problem, where we are given the
graph shown in \autoref{fig:edge_exe_graph}, whose node attributes are listed
\autoref{tab:edge_exe}. The first two attributes are categorical, denoting the company employing
each user, either \textsf{G} or \textsf{F}. Instead of a classic binary one-hot encoding, we
additionally use $-1$ to denote that a user had been employed by a given company. \textsf{Gender} is
encoded as presented earlier, and we also apply a pre processing to transform \textsf{age} between
$-1$ and $1$. Finally, \textsf{opinion} represents the view of former \textsf{F} employees about a
societal issue, with $0$ denoting a neutral position. In \autoref{tab:edge_exe}, we also show two
directions that explain the connections in the graph $G$. The first covers being of the same age and
same companies, especially at \textsf{G}, while the second directions covers former employees of
\textsf{F} having different \textsf{age} and \textsf{opinion}. Not only are the results
interpretable as claimed before, but we can also assess the strength of each explanation, using
their goodness. For instance, we see that the edge between $x$ and $y$ has a low score, because
those two users have indeed the same \textsf{age} and \textsf{opinion}.

\begin{figure*}[htpb]
  % \centering
  \begin{subfigure}[t]{0.55\textwidth}
    \centering
    \includegraphics[width=.9\textwidth]{tikz/edge_example_tikz.pdf}
    \caption{The graph $G$. Larger nodes represent old users while smaller ones are younger. Their
    gender is inscribed in the node and their shape represent their opinion: rectangle for $-1$,
    circle for $0$ and diamond for $+1$. Finally, orange stands for being employed at company G and
    faded blue for having been employed at company F. As for the edges, the red ones are labeled by
    the first direction and the green ones by the second direction, and we label them with the
    associated (normalized) goodness. } \label{fig:edge_exe_graph}
  \end{subfigure}~
  \begin{subfigure}[t]{0.42\textwidth}
    \centering
    \small
    \begin{tabular}{lrrrrr}
      \toprule
      {}    & G   & F    & gender & age    & opinion \\
      \midrule
      $u$   &  $+1$ &   $0$ &   $-1$ &  $-0.8$ &     $0$ \\
      $v$   &  $+1$ &  $-1$ &   $-1$ &  $-0.7$ &    $+1$ \\
      $w$   &  $+1$ &   $0$ &   $+1$ &   $0.6$ &    $+1$ \\
      $x$   &  $+1$ &  $-1$ &   $+1$ &   $0.9$ &    $-1$ \\
      $y$   &   $0$ &  $-1$ &   $-1$ &   $0.7$ &    $-1$ \\
      $z$   &   $0$ &  $-1$ &   $+1$ &  $-0.6$ &    $+1$ \\
      \midrule
      $w_1$ &  $+3$ &  $+1$ &    $0$ &    $+1$ &     $0$ \\
      $w_2$ &  $0$  &  $+4$ &    $0$ &    $-2$ &    $-4$ \\
      \bottomrule
    \end{tabular}
    \caption{The numerical attributes of the six nodes. They are followed by the (non normalized)
    two directions $w_1$ and $w_2$.} \label{tab:edge_exe}
  \end{subfigure}
  \caption{A small instance of \ecp{} and an handcrafted solution, albeit non-optimal.}
  \label{fig:edge_exe}
\end{figure*}

\paragraph{Choice of $k$}
\label{par:edge_choose_k}

For clarity, the previous example of \autoref{fig:edge_exe} has only a handful of nodes and
dimensions, and can therefore reasonably be explained by two directions. But in larger graphs, the
choice of $k$ is a legitimate question.
At one extreme, picking $k=|E|$ renders the topology of the graph
irrelevant, as we can simply exploit the trivial
solution of having a single direction $w_{uv} = \frac{s_{uv}}{||s_{uv}||}$ for every edge $(u,v)$.
More generally, as $k$ increases, we expect the value $\max_{\mathcal{D}_k} \sum_{u,v \in E}
\max_{\ell \in \rangesk} g(s_{uv}, w_{\ell})$ to increase as
well, at the cost of interpretability. There could be a principled approach to finding the
\enquote{best} $k$, based on a information theoretical measure of the complexity of $\mathcal{D}_k$
and the minimum description length principle~\autocite{grunwald2005tutorial}. However, in the
interest of simplicity, in the following we focus on the formulation stated in
\autoref{p:edge_full}, where $k$ is given. In practice, we further constrain $k$ to be small (say
less than \np{10}). Within the multilayer framework, $k$ could also be seen as the number of
underlying layers. To retain a fair level of interpretability, we deem appropriate to have $k$ not
too large.

\paragraph{Topological constraints}

In addition to assume that the input graph is the superposition of a small number of layers,
another way to leverage the graph topology is to define local constraints at the node level.
For that we will consider node profiles to be normalized and having unit norm. A practical
justification is that if nodes are
users of a social network and the profiles measure their activity across several domains, this
allows to compare users with various level of total activity. Furthermore, we assume that profiles
are linear combination of directions, that
is $x_u = b_u + \sum_{v \in \nei(u)} a_{uv} w_{uv}$, where $a_{uv}$ are real weights and $b_u
\in \Rbb^d$ is the bias vector of $u$, its inherent profile take in isolation.
Viewed in the other direction, this can be interpreted as follow: each time a node $u$
connect to another node $v$ along a direction $w_{uv}$, it \enquote{consumes} a part of its profile
proportional to $w_{uv}$. Therefore, we subtract the term
\begin{equation}
  \label{eq:edge_node_loss}
  \mathcal{L}_{\mathrm{node}} =
  \sum_{u\in V} \left|\left| x_u - b_u - \sum_{v \in \nei(u)} a_{uv} w_{uv} \right|\right|^2_2
\end{equation}
from \eqref{eq:edge_full}.

We also make the assumption that each node $u$ is only involved in $k_\mathrm{local} < k$
different directions. The intuition is that when $d$ is large enough, users have to focus their
energy and can only express their interest in a few number of dimensions. In other words, all the
edges incident to $u$ can only be associated with one of $k_\mathrm{local}$ directions. For
some values of
$k_\mathrm{local}$, this is a \NPc{} graph colouring problem and thus cannot necessarily be enforced
exactly. Still, it can be seen as a learning bias and will be used to generate synthetic data.
Furthermore, we can also express it as an optimization constraint. Specifically, suppose that for
an edge $(u,v)$ associated with the direction $w_i \in \mathcal{D}_k$, we have a $k$-dimensional
vector $y_{uv}$ equal to $e_i$ (that is all zero except for the \ith{} component set to one). Let
us stack all such vectors into a $|E| \times k$ matrix $Y$, and let $C \in \Rbb^{|V| \times |E|}$ be
the incidence matrix of $G$\footnote{The incidence matrix of a graph $G$, usually denoted by $B$, is
the $|V| \times |E|$ matrix such that $B_{i,j} = 1$ if the node $v_i$ and edge $e_j$ are incident
and 0 otherwise.}, with $c_u$ being its $u^{th}$ row. Then $c_u^TY$ is a $k$-dimensional vector that counts
how many times each direction is incident to node $u$. We thus want the $\ell_0$ norm of $c_u^TY$
(\ie the number of its non-zero component) to be upper bounded by $k_\mathrm{local}$.
This is not a convex constraint but we can relax it, for such sparsity inducing problems have been
well studied~\autocite{sparseOptim12}. Namely, we let $y_{uv}$ be a softmax membership
$k$-dimensional vector, that is we set $\forall i \in \rangesk\, y_{uv,i} =
\frac{\exp(\beta{s_{uv}}^T w_i)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)} $, with $\beta>0$ a large
positive constant.  And we replace the $\ell_0$ norm by a $\ell_1$ norm (or even a $k$-support
norm~\autocite{KsupportNorm12}), yielding the following $\mathcal{L}_{\mathrm{local}}$ to be
minimized:
\begin{equation}
  \label{eq:edge_local_loss}
  \mathcal{L}_{\mathrm{local}} =
  \sum_{u \in V} \left\| \sum_{v \in \nei(u)} y_{uv} \right\|_1 =
  \sum_{u \in V} \sum_{\ell = 1}^k \sum_{v \in \nei(u)}
  \frac{\exp(\beta{s_{uv}}^T w_\ell)}{\sum_{j=1}^k \exp(\beta {s_{uv}}^T w_j)}
\end{equation}

These two constraints can be combined, so that the profiles are close to a linear combination of a
small number of directions.

@@

Here we show why \autoref{p:edge_full} can model signed graphs as a special case, namely with $k=2$
and specifically crafted profiles.
Remember the learning
bias we introduced in the previous chapter \vpageref{text:cc_new_bias}: every node belongs to
exactly one of the $K$ clusters available, is connected positively with the other nodes belonging to
the same cluster, and is connected negatively with the nodes belonging to every other clusters. Now
let the dimension of profiles be $d=K$ and the number of directions $k=2$. Moreover, we define the
two direction vectors as $w_1 = \frac{\onev}{\sqrt{d}} = w_+$ and $w_2 = -w_+ = w_-$, where
$\onev$ is the vector of dimension $d$ with only $1$ coordinates. Finally, for a node $u$ in
cluster $i$, we assume its profile $x_u$ is set as follow: all component are equal to $-b$ except
for the \ith{} one, which is equal to $a$, where $a$ and $b$ are real constants we define now.

First, because $x_u$ is a unit vector, we have that $a^2 + (K-1)b^2=1$, from which $a =
\sqrt{1-(K-1)b^2}$ whenever $1-(K-1)b^2 \geq 0$, or equivalently $b \leq \frac{1}{\sqrt{K-1}}$.
Next, we look at $\left(x_u \circ x_v\right)^T \onev = {s_{uv}}^T \onev = \frac{1}{\sqrt{d}}
\sum_{i=1}^d  {s_{uv}}_{;i}$ for any edge $(u,v) \in E$. If $u$ and $v$ are in the same cluster, then
$\sum_{i=1}^d  {s_{uv}}_{;i} = a^2 + (K-1)b^2 = 1$, implying that ${s_{uv}}^T w_+ = \frac{1}{d} >
{s_{uv}}^T w_- = -\frac{1}{d}$. Otherwise, $\sum_{i=1}^d  {s_{uv}}_{;i} = (K-2)b^2 - 2ab = (K-2)b^2 -
2b\sqrt{1-(K-1)b^2} = f_K(b)$. By computing the derivative of $f_K$ with respect to $b$, we find
that it reaches a minimum of $-\frac{1}{K-1}$ at $b = \frac{1}{\sqrt{K(K-1)}} \leq
\frac{1}{\sqrt{K-1}}$. In this case, ${s_{uv}}^T w_+ = -\frac{1}{(K-1) \sqrt{d}} < {s_{uv}}^T w_- =
\frac{1}{(K-1) \sqrt{d}}$. To maximize $\sum_{(u,v)\in E} \max_{\ell \in \{1, 2\}} {s_{uv}}^T
w_\ell$, we therefore assign $w_+$ to edges whose both endpoints belong to the same cluster, and
$w_-$ to edges joining two different clusters.

@@

@@

A simple and natural baseline is to cluster the set of all similarity edge vectors $
\theset{s_{uv}}{(u,v) \in E}$ using the $k$-means algorithm, where $k$ is the number of directions
set in \autoref{p:edge_full}. Formally, given the graph $G=(V,E)$ and the node profiles $X$ as
input, we order the edges from $e_1$ to $e_m$. Then we build the matrix $S \in \Rbb^{m\times d}$, whose
\ith{} row is the similarity $s_{uv} = x_u \circ x_v$ between the profiles of the endpoints of $e_i=(u,v)$.
Running $k$-means on the rows of $S$ will thus partition $E$ into $k$ sets of edges, call them $E_1,
\ldots, E_k$. A drawback of that simplicity is that it does not solve \autoref{p:edge_full}.
Indeed, the partition of $E$ does not provide the set of directions $\mathcal{D}_k = \{w_1, w_2,
\ldots, w_k\} \in \dsphere$ we are looking for. A natural way of obtaining directions from that
partition is to set $w_\ell$ to be the normalized cluster center of $E_\ell$. We refer to this
method as \kmeans{}, and assuming it requires $T$ iterations to converge, it has a linear complexity
of $O(T|E|)$.
\iffalse
\begin{center}
  \rule{\textwidth}{.3pt}
  \begin{algorithmic}[1]
    \Function{\kmeans{}}{graph $G=(V,E=\{e_1, \ldots, e_m \})$, profiles $X$}
    \State build the matrix $S$ such that $S_{i,j} = {s_{e_i}}_{;j}$
    \State let $\yhat{}$ and $\{c_1, \ldots, c_k\}$ be the assignment and centroids resulting of
    running $k$-means on the rows of $S$
    \State \textbf{return} \yhat{}, $\mathcal{D}_k = \left\{\frac{c_\ell}{\left\| c_\ell
    \right\|}\right\}_{\ell=1}^k $
    \EndFunction
  \end{algorithmic}
  \rule{\textwidth}{.3pt}
\end{center}
\fi

However,
this does not guarantee that we have assigned the best direction to every edge. We therefore propose
the following heuristic, inspired by the Lloyd algorithm for $k$-means, but where the directions in
$\mathcal{D}_k$ play the role of centroids. Namely, we alternate between two steps:
\begin{enumerate}[(i), nosep]
  \item create a new partition $\{E'_1, \ldots, E'_k\}$ by assigning every edge $(u,v)$ to the
    direction $w$ maximizing its goodness ${s_{uv}}^T w$, that is $E'_\ell = \theset{(u,v) \in E}{
    w_\ell = \argmax_{w_i \in \mathcal{D}_k} {s_{uv}}^T w_i}$, and
  \item update the directions to make them optimal with respect to the edges currently assigned to
    them. Specifically, set $\mathcal{D}_k = \left\{ w_\ell = \frac{s_\ell}{\left\| s_\ell
    \right\|}\right\}_{\ell=1}^k $, where $s_\ell = \sum_{u,v \in E'_\ell} s_{uv}$ is the sum of all
    the similarity vectors of the edge in $E'_\ell$,
\end{enumerate}
until the edge assignment is stable or we reach a maximum number of iterations. By analogy with the
$k$-means algorithm, obtaining convergence guarantees in the general case would be surprising, but in
practice we expect that only a small number $T$ of iterations would be needed. We call this method
\lloyd{} and note it also has a linear complexity of $O(T|E|)$.

\subsection{Convex relaxation}
\label{sub:edge_vector}
